A Dearth of Young Stars and Planets in the Kepler Field

### 2024-06-05 paper/ms.tex
update ms

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of these rotators are
  suitable for gyrochronology (dark gray points), based on factors
  including their non-binarity and proximity to the main
  sequence (see Section~\ref{subsec:flags}).  Some
  are ``confirmed'' or ``candidate'' KOIs that meet an additional set
  of planetary quality criteria (yellow points;
  Section~\ref{subsec:plflags}).  Surface gravities and effective
  temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

To assess the statistical uncertainties of these rotation periods, we
compared our adopted \citetalias{Santos_2019} and
\citetalias{Santos_2021} periods with those reported by
\citet{McQuillan_2014} (see Appendix~\ref{app:reinhold23}).  We found
that for $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a
precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of
$P_{\rm rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the
period difference increased roughly linearly with period.  Based on
this comparison, we adopted a simple prescription for the period
uncertainties, such that there are 1\% relative uncertainties below
$P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods of
30 days.

@@

\section{Ages From Lithium}
\label{sec:liage} 
\subsection{Sample Summary}
\label{subsec:tally}

The rotation periods for the star and planet samples are shown in
Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods and a reliable
DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
these stars to have temperatures of 3800--6200\,K, yields
\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
apparently single, near the main sequence, with $\log g$$>$4.2,
$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
\nuniqstarsantosallbutruwe\ rotators potentially amenable for
gyrochronology.  Finally, imposing RUWE$<$1.4 and requiring no
possible blends ($\Delta G$<2.5) within 4$''$ leaves
\nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
to be valid.

In the planet sample, the cumulative KOI table included
\nnonfpcumkois\ confirmed and candidate planets, orbiting
\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
cuts, save for that on RUWE.  We can then require these confirmed and
candidate planets to have MES$>$10.  This combination of selection
criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
rotation-based ages orbiting
\nplhoststarwgyroagewithgrazingandhighruwe\ stars
(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
If we additionally require the planets to be non-grazing ($b<0.8$), we
are left with \nplwgyroagenograzing\ planets orbiting
\nplhoststarwgyroagenograzing\ stars.

@@

To calculate lithium ages, we use \texttt{EAGLES} (git commit
\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
a purely empirical interpolation approach.  \texttt{EAGLES} 
was calibrated on lithium measurements of stars observed by the
Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
uncertainties for purposes of interfacing with \texttt{EAGLES} by
taking the maximum of the measured positive and negative 1$\sigma$
uncertainty intervals.

@@

@@

Let us focus discussion on blemish-free stars with 
confirmed planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must also have no quality flags
raised.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data.  We derived new ages using empirical
interpolation-based methods, and assessed the reliability of these ages
by comparing them against benchmark open clusters
(Figure~\ref{fig:agescalecompone}).

Our analysis blindly recovered the ages of all 14 known Kepler planets
in clusters.  While lithium provided only marginal added information
for most of the sample, the lithium and rotation ages agreed in
$\gtrsim$90\% of cases for which comparison was possible.
Our resulting ages included two-sided $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ sub-gigayear confirmed planets, and
two-sided $t_{\rm gyro}$ with one-sided $t_{\rm Li}$ for
\ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
systems with RUWE$>$1.4 and including ``candidate'' planets
would increase the counts to
\ltonegyrmediumqconfirmedtwosided\ and
\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
such as on transit geometry, would further increase the counts.  The sizes and
periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, and others include Earth-sized 
planets a few hundred million years old.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item The age distributions of both the Kepler target stars and the
    known Kepler planet-hosts show a demographic cliff.  There are
    twice as many ``old'' (2-3\,Gyr) stars in the
    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
    three billion years ago.
    This result from rotation-based ages broadly agrees with recent reports of
    a declining star-formation rate
    from CMD fitting and white-dwarf chronology, though with
    important nuances.
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the
    Kepler field implies that we should instead expect $\approx$250.
  %
  %FIXME: 120 number
  \item Using rotation-based ages, we have derived ages for
    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
    and \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$120-170 known sub-gigayear planets with well-constrained
    ages.  This lessens the original factor of ten discrepancy to
    a factor of at most two.
\end{enumerate}

@@

\section{The Preferred Rotation Period Catalog}
\label{app:reinhold23}

@@

While we opted for the \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation period catalogs, the more recent
analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
used a similar selection function as \citeauthor{Santos_2021}
(i.e.~all Kepler stars), and considered a novel period measurement
approach based on gradients of power spectra (GPS).  Their method
likely provides greater completeness, due to its greater sensitivity
to signals that are only weakly periodic.

>>>>

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of these rotators are
  suitable for gyrochronology (dark gray points), based on factors
  including their non-binarity and proximity to the main
  sequence (see Section~\ref{subsec:flags}).  Some
  host ``confirmed'' or ``candidate'' KOIs that meet additional
  planetary quality criteria (yellow points;
  Section~\ref{subsec:plflags}).  Surface gravities and effective
  temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

To assess the statistical uncertainties of these rotation periods, we
compared our adopted periods with those reported by
\citet{McQuillan_2014}.  The details are in Appendix~\ref{app:reinhold23}.  We found
that for $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a
precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of
$P_{\rm rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the
period difference increased roughly linearly with period.  Based on
this comparison, we adopted a simple prescription for the period
uncertainties, such that there are 1\% relative uncertainties below
$P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods of
30 days.

@@

\section{Ages From Lithium}
\label{sec:liage}

@@

To calculate lithium ages, we use \texttt{EAGLES} (git commit
\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
an empirical interpolation approach.  \texttt{EAGLES} 
was calibrated on lithium measurements of stars observed by the
Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
uncertainties for purposes of interfacing with \texttt{EAGLES} by
taking the maximum of the measured positive and negative 1$\sigma$
uncertainty intervals.

@@

%\subsection{Sample Summary}
%\label{subsec:tally}
%
%The rotation periods for the star and planet samples are shown in
%Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
%observed by Kepler, requiring reported rotation periods and a reliable
%DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
%these stars to have temperatures of 3800--6200\,K yields
%\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
%apparently single, near the main sequence, with $\log g$$>$4.2,
%$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
%\nuniqstarsantosallbutruwe\ rotators.
%Finally, imposing the full set of quality cuts,
%i.e.~further requiring RUWE$<$1.4 and requiring no
%resolved blends ($\Delta G$<2.5) within 4$''$ leaves
%\nuniqstarsantosrotgyroappl\ stars for which we expect rotation-based
%age dating to be secure.
%
%
%In the planet sample, the cumulative KOI table included
%\nnonfpcumkois\ confirmed and candidate planets, orbiting
%\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
%cuts, save for that on RUWE.  We can then require these confirmed and
%candidate planets to have MES$>$10.  This combination of selection
%criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
%rotation-based ages orbiting
%\nplhoststarwgyroagewithgrazingandhighruwe\ stars
%(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
%If we additionally require the planets to be non-grazing ($b<0.8$), we
%are left with \nplwgyroagenograzing\ planets orbiting
%\nplhoststarwgyroagenograzing\ stars with rotatio-based ages.
%This cut drops $\approx$41 more planets than one would expect if the initial sample
%of \nplwgyroagewithgrazingandhighruwe\ planets had zero false positives.
%The binomial probability that this could happen by coincidence is 0.1\%.

@@

Let us focus discussion on blemish-free stars with 
confirmed planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must also have no quality flags
raised.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.
%FIXME TODO GET MORE PRECISE LANGUAGE -- SYSTEMS OR PLANETS OR STARS?

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data from the Kepler field.  We derived new ages using empirical
interpolation-based methods, and assessed the reliability of these ages
by comparing them against benchmark open clusters
(Figure~\ref{fig:agescalecompone}).

Our analysis blindly recovered the ages of all 14 known Kepler planets
in clusters.  While lithium provided only marginal added information
for most of the sample, the lithium and rotation ages agreed in
$\gtrsim$90\% of cases for which comparison was possible.
Our resulting ages included two-sided $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ sub-gigayear confirmed planets, and
two-sided $t_{\rm gyro}$ with one-sided $t_{\rm Li}$ for
\ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
systems with RUWE$>$1.4 and including ``candidate'' planets
would increase the counts to
\ltonegyrmediumqconfirmedtwosided\ and
\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
such as on transit geometry, would further increase the counts.  The sizes and
periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an evolutionary
connection to their youth.
Other discoveries include Earth-sized 
planets with new ages of only a few hundred million years.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item The age distributions of both the Kepler target stars and the
    known Kepler planet-hosts show a demographic cliff.  There are
    twice as many ``old'' (2-3\,Gyr) stars in the
    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
    three billion years ago.
    This result from rotation-based ages broadly agrees with recent reports of
    a declining star-formation rate
    from CMD fitting and white-dwarf chronology, though with
    some nuances.
    One important advance is that our
    rotation-based ages offer better accuracy at $t$$<$1\,Gyr
    than current isochronal methods (Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the
    Kepler field implies that we should instead expect $\approx$250.
  %
  %FIXME: 120 number
  \item We have derived rotation-based ages for
    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
    and \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$120-170 known sub-gigayear planets with well-constrained
    ages.  This lessens the original factor of ten discrepancy to
    a factor of at most two.
\end{enumerate}

@@

\section{Rotation Period Catalog Comparison}
\label{app:reinhold23}

@@

While we opted for the \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation period catalogs, the recent
analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
used a similar selection function as \citeauthor{Santos_2021}
(all Kepler stars), and considered a novel period measurement
approach based on the gradient of the power spectrum (GPS).  Their method
likely provides greater completeness, due to its greater sensitivity
to signals that are only weakly periodic.
